Reassessing the role of theory and machine learning in forecasting civil conflict

### 2020-09-20 paper.Rmd
third shot

B\&S advocate the use of theory to guide prediction. But "theory" as used by B\&S is an ambiguous and undefined concept. `Theory'' has many interpretations in social science.  A theory is a system of internally consistent, logically connected propositions from which additional statements may be deduced.  The additional statements can be subjected to empirical scrutiny.  An example might be William Riker's theory of coalitions which posits propositions in game theoretic form from which the size principal of minimal winning coalitions may be deduced and further examined empirically.  For the purposes of discussion we call this strong theory.

However, this is not what many mean by theory in the contemporary social science. For many,
a theory is a falsifiable story or narrative that makes causal claims about the relationships among variables. A theory in this sense could be entirely created by the scholar. More typically it is fused together from a meta-analysis of prior work on the topic under consideration.   We call these types of theory weak theory.

They are weak because they do not include two characteristics of strong theory, though they may include one:
1. A theory is separate from the case or cases being examined. It is intented to be general.
2. A theory generates new statements that in principle can be empirically falsified, not proven.

Most social science research uses a weak form of theory.  In the case of B\&S they base their theoretical thinking on research that argues that conflict is not the product of structural factors, but rather comes from an interaction of state repression and the changes in costs and benefits to citizens for engaging in anti-government conflict. Following a wide swath of scholars, they argue that the escalation and de-escalations of interactions between citizens and government are what typically lead up to the onset of civil wars.  Operationally, this boils down to four types of concepts: demands, accommodations,
nonviolent repression, and low-level violence.
\begin{itemize}
\item Demands include events in which either opposition or rebels is the source and
government is the target and typically involve appeals for political or institutional
reform, policy change, or extension of rights.
\item Nonviolent repression and accommodation include events in which government
is the source and either opposition or rebels is the target. Nonviolent
repression includes threats, sanctions, restrictions on freedom of movement, or
rejections of group demands.
\item Low-level violence includes all events in which
(1) government is the source and opposition is the target, (2) opposition is the
source and government is the target, (3) government is the source and rebels are
the target, or (4) rebels are the source and government is the target.
\end{itemize}
These variables, taken from ICEWS, are the main components of the escalation model in the B\&S article.

Indeed most empirical social science research follows this approach: 1) delineation of an argument, 2) selection of variables that represent a plausible or compelling representation of the concepts, followed by 3) utilization of these variables in a statistical model with a dependent variable that measures the main phenomena of investigation. It is worth pointing out that this is simply one set of variables that are selected to represent the concepts in the theoretical story. A strong theory gives more guidance on compelling choices than does a weak theory.  Accordingly, while the choice is plausible is not necessarily compelling.  Another choice might be equally plausible.  In the ICEWS example, we might include variables on intragovernmental demands (from the military toward the central government), or further look a the dynamics among opposition groups.    In practical terms there is a wide array of choices that are available to instantiate any particular theory.  A second point is that putting them in a linear regression may be a particularly crude instantiation of complicated temporal dynamics.  One general principal that is on point in the B\&S article and is more widely relevant is that dynamical phenomenon like civil wars need to have explanatory models that change rapidly over time---and by implication that are represented by data that change fairly quickly.

MORE HERE ON GENERAL IDEAS.  ANDY ANY BULLET pOINTs TO ADD?

Another important consideration when developing a predictive model is to do so without contaminating the out-of-sample test set. This occurs if test set performance is used to inform specification---predictive modeling's analogue to p-hacking---and special care is needed [e.g. see the approach by @colaresi:mahmood:2017]. B&S did pre-register their 2016-H1 forecasts, which verifies the integrity of the forecasts, but this does not assure the model development process. To be clear: there is no reason to suspect test set contamination in the escalation or other models. Rather, it is challenging to provide verifiable assurance thereof in a single analysis iteration that both develops and tests a model on data that was available contemporaneously.

Moreover, they misunderstand the use of ICEWS event data in current research. They claim that most applications to date have focused on the quad categories, but this ignores a wide swath of literature [@steinert-threlkeld:2017; and @metternich:etal:2013] that uses a specific action---such as protest---defined in the CAMEO ontology. In their article on conflict in Thailand, Metternich et alia hand-coded, for example, every actor in Thailand and focused on an analysis of how those interact.

We encountered several issues in the code underlying the B&S analysis. The problems we encountered are not subjective modeling choices. When we fix these issues and perform an updated analysis, the B&S conclusions are all essentially overturned. In other words, B&S findings are based on a faulty analysis, invalid, and should be disregarded.

@@

Blair & Sambanis have focused attention on comparing forecasting models and forecasts in the civil war domain.  As social science becomes more adept at predictive analysis, this will doubtless be of increasing importance.  However, these comparisons must be made carefully to ensure those correct inferences are drawn.  We continue to think that theory is overrated [@ward:2016], and that machine learning and big data will allow us to learn new things. It will be interesting to see how the evidence for these claims is adjudicated with additional usage and careful evaluation.

>>>>

B\&S advocate the use of theory to guide prediction. But "theory" is not defined by B\&S, and may an ambiguous and undefined concept. `Theory'' has many interpretations in social science.  For some---the EITM movement, for example---theory is a system of internally consistent, logically connected propositions from which additional statements may be deduced.  The additional statements can be subjected to empirical scrutiny.  An example might be William Riker's theory of coalitions which posits propositions in game theoretic form from which the size principal of minimal winning coalitions may be deduced and further examined empirically. A more recent example is found in The Political Logic of Survival.  For the purposes of discussion we call this strong theory.

However, for many,
a theory is a falsifiable story or narrative that makes causal claims about the relationships among variables.  Typically it is fused together from a meta-analysis of prior work on the topic under consideration.   We call these types of theory weak theory.
They are weak because they do not include two characteristics of strong theory, though they may include one:
1. A theory is separate from the case or cases being examined. It is intented to be general.
2. A theory generates new statements that in principle can be empirically falsified, not proven.

Most social science research uses a weak form of theory.  In the case of B\&S they base their theoretical thinking on research that argues that conflict is not the product of structural factors, but rather comes from an interaction of state repression and the changes in costs and benefits to citizens for engaging in anti-government conflict. Following a wide swath of scholars, they argue that the escalation and de-escalations of interactions between citizens and government are what typically lead up to the onset of civil wars.  Operationally, this boils down to four types of concepts: demands, accommodations,
nonviolent repression, and low-level violence.
\begin{itemize}
\item Demands include events in which either opposition or rebels is the source and
government is the target and typically involve appeals for political or institutional
reform, policy change, or extension of rights.
\item Nonviolent repression and accommodation include events in which government
is the source and either opposition or rebels is the target. Nonviolent
repression includes threats, sanctions, restrictions on freedom of movement, or
rejections of group demands.
\item Low-level violence includes all events in which
(1) government is the source and opposition is the target, (2) opposition is the
source and government is the target, (3) government is the source and rebels are
the target, or (4) rebels are the source and government is the target.
\end{itemize}
These variables, taken from ICEWS, are the main components of the escalation model in the B\&S article. The escalation theory is that the forces behind these variables interact such that as each of them increases, they independently increase the probability of civil conflict.  The actual mechanisms of the dynamism among these variables are not specified.

Indeed most empirical social science research follows this approach: 1) delineation of an argument, 2) selection of variables that represent a plausible or compelling representation of the concepts, followed by 3) utilization of these variables in a statistical model with a dependent variable that measures the main phenomena of investigation. It is worth pointing out that this is simply one set of variables that are selected to represent the concepts in the theoretical story. A strong theory gives more guidance on compelling choices than does a weak theory.  Accordingly, while the choice is plausible is not necessarily compelling.  Another choice might be equally plausible.  In the ICEWS example, we might include variables on intragovernmental demands (from the military toward the central government), or further look a the dynamics among opposition groups.    In practical terms there is a wide array of choices that are available to instantiate any particular theory.  A second point is that putting them in a linear regression may be a particularly crude instantiation of complicated temporal dynamics.  One general principal that is on point in the B\&S article and is more widely relevant is that dynamical phenomenon like civil wars need to have explanatory models that change rapidly over time---and by implication that are represented by data that change fairly quickly.
At this level of detail the theoretical model is more about chosing the correct variables.  And while we don't undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.

Most importantly, a single study is unlikely to be convincing that theory-based models are better (or worse) than ad hoc models based mainly in machine learning.  Even if B\&S were correct that in their study the theory based escalation model performed better, it would simple be a case study of one case where it was preferable. It establishes no general tendency, let alone law-like evidence for general theory.  Nor, against it.

#An important consideration when developing a predictive model is to do so without contaminating the out-of-sample test set. This occurs if test set performance is used to inform specification---predictive modeling's analogue to p-hacking---and special care is needed [e.g. see the approach by @colaresi:mahmood:2017]. B&S did pre-register their 2016-H1 forecasts, which verifies the integrity of the forecasts, but this does not assure the model development process. To be clear: there is no reason to suspect test set contamination in the escalation or other models. Rather, it is challenging to provide verifiable assurance thereof in a single analysis iteration that both develops and tests a model on data that was available contemporaneously.

#Moreover, they misunderstand the use of ICEWS event data in current research. They claim that most applications to date have focused on the quad categories, but this ignores a wide swath of literature [@steinert-threlkeld:2017; and @metternich:etal:2013] that uses a specific action---such as protest---defined in the CAMEO ontology. In their article on conflict in Thailand, Metternich et alia hand-coded, for example, every actor in Thailand and focused on an analysis of how those interact.

We encountered several issues in the code underlying the B&S analysis. The problems we encountered are not subjective modeling choices.  They bear no implications on the question of whether theory based model are preferable.  However,  when we fix these issues and perform an updated analysis, the B&S conclusions are all essentially overturned. In other words, B&S findings are based on a faulty analysis, invalid, and should be disregarded as evidence in favor of theory-based models.

@@

Blair & Sambanis have focused attention on comparing forecasting models and forecasts in the civil war domain.  As social science becomes more adept at predictive analysis, this will doubtless be of increasing importance.  However, these comparisons must be made carefully to ensure those correct inferences are drawn.  We continue to think that theory is overrated [@ward:2016], and that machine learning and big data will allow us to learn new things. It will be interesting to see how the evidence for these claims is adjudicated with additional usage and careful evaluation.  Social science has leaned on theory for (at least) seventy years, but it might be time to find useful ways to incorporate some of the recently developed methods in the arena of machine learning as well.
